Refactoring \texttt{nls()}: a constructive critique

### 2023-01-31 RefactoringNLS/RefactoringNLS.tex
note on profiling

  pdftitle={Refactoring the nls() function in R},

@@

\title{Refactoring the \texttt{nls()} function in R}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2022-6-15}

@@

Important considerations are the usability and maintainability of the
code base that provides the functionality \texttt{nls()} claims to
offer. Our work suggests that the existing code makes maintenance and
improvement very difficult, with legacy applications and examples
blocking some important updates. Discussion of these matters is relevant
to improving R generally, as well as its nonlinear estimation tools.

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones which call
  calculations in other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to a number of pre-coded (``selfStart'') models
\end{itemize}

@@

Adding a small quantity to the loss function before dividing avoids
trouble. In 2021, one of us (J. Nash) proposed that
\texttt{nls.control()} have an additional parameter \texttt{scaleOffset}
with a default value of zero. Setting it to a small number -- 1.0 is a
reasonable choice -- allows small-residual problems (i.e., near-exact
fits) to be dealt with easily. We call this the \textbf{safeguarded
relative offset convergence criterion}. The default value gives the
legacy behaviour. We are pleased to note that this improvement is now in
the R distributed code as of version 4.1.0.

@@

The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from
analytic expressions, while that of \texttt{nls()} mostly uses a
numerical approximation. Some selfStart models do provide ``gradient''
information. \texttt{minpack.lm::nlsLM()} invokes
\texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
it appears to use an internal approximate Jacobian code from the
original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such
differences in approach can lead to different behaviour, though in our
experience differences are usually minor.

@@

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file and in package
\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
may provide the Jacobian in the ``gradient'' attribute of the
``one-sided'' formula that defines each model, and these Jacobians may
be the analytic forms. The \texttt{nls()} function, after computing the
``right hand side'' or \texttt{rhs} of the residual, then checks to see
if the ``gradient'' attribute is defined, and, if not, uses
\texttt{numericDeriv} to compute a Jacobian into that attribute. This
code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
functions. The use of analytic Jacobians almost certainly contributes to
the good performance of \texttt{nls()} on selfStart models.

@@

Our concern with \texttt{na.action} is that users may be unaware of the
effects of an option setting they may not even be aware has been set. Is
\texttt{na.fail} a safer default?

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates outside of \texttt{nls()}, in
part because it may call that function. Such circular references are, in
our view, dangerous. Moreover, we believe that it would be helpful to
have selfStart models that allow users to explicitly provide values
other than those suggested for the starting parameters.

Other concerns are revealed by the example below. Here, the
\texttt{SSlogis} selfStart function is used to generate a set of initial
parameters for a 3-parameter logistic curve. The form used by
\texttt{SSlogis} is
\(y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))\), but we
show how the starting parameters for this model can be transformed to
those of another form of the model, namely,
\(y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))\).

@@

In the event that a selfStart model is not available, \texttt{nls()}
sets all the starting parameters to 1. This is, in our view, tolerable,
but could be improved by using a set of values that are all slightly
different, which, in the case of the example model
\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
avoid a singular Jacobian because \(b\) and \(d\) were equal in value. A
sequence like 1.0, 1.1, 1.2, 1.3 for the four parameters could be
provided quite simply instead of all 1's.

\hypertarget{strategic-issues-in-selfstart-models}{%
\subsubsection{Strategic issues in selfStart
models}\label{strategic-issues-in-selfstart-models}}

By providing starting values that are likely to be reasonable and by
including Jacobian code in the selfStart model code, some of the
deficiencies of the Gauss-Newton algorithm are mitigated. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

Let us compare the \texttt{nls()} result with that from
\texttt{nlsr::nlxb()}, with ostensibly solves the same problem:

@@

The result object from \texttt{nlsr::nlxb()} is still much smaller than
the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
returns the sum of squares as well as the residual vector and Jacobian.
The counts of evaluations are also returned. (Note that the singular
values of the Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result.) As we made our comparisons, We
noted several potential updates to the \texttt{nlsr} documentation as
well as that for \texttt{nls()}.

@@

The variable projection method (Golub and Pereyra
(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
effective then general approaches in finding good solutions to nonlinear
least squares problems when some of the parameters appear linearly.
However, setting up the calculations, that is, identifying which
parameters are linear, is not a trivial task.

@@

The nonlinear modeling model specifications are, of course, a
development of linear ones. Unfortunately, the explicit model
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
linear modeling function \texttt{lm()}, which requires this model to be
specified as \texttt{y\ \textasciitilde{}\ x}.

@@

  \DocumentationTok{\#\# Then we use nls\textquotesingle{} indexing feature for parameters in non{-}linear}
  \DocumentationTok{\#\# models to use the conventional algorithm to fit a model in which}
  \DocumentationTok{\#\# alpha and beta vary with Strip.  The starting values are provided}
  \DocumentationTok{\#\# by the previously fitted model.}
  \DocumentationTok{\#\# Note that with indexed parameters, the starting values must be}
  \DocumentationTok{\#\# given in a list (with names):}
  \DocumentationTok{\#\# ?? but why use b here AND in the new formula??}
\NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
\NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{2}\NormalTok{], nl), }\AttributeTok{b =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{3}\NormalTok{], nl), }\AttributeTok{th =}\NormalTok{ b[}\DecValTok{1}\NormalTok{]))}
  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.2}\NormalTok{)}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Indexed models clearly have a place in some areas of research. We need
to be able to process models such as
\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)}
where \texttt{Strip} is the index. Data \texttt{Length} and
\texttt{Conc} are available for all observations and parameter
\texttt{th} applies to every fitted point, but \texttt{a{[}Strip{]}} and
\texttt{b{[}Strip{]}} apply only to data indexed, which could be an
experimental run number, or some other categorization.

Given that \texttt{{[}\ {]}} are the R method to index arrays, their
presence in a model formula signals indexing. The question is how to
perform the correct calculations.

@@

We anticipate that tools like \texttt{nlsr} will become more attractive
as users discover the increased robustness and general capabilities,
along with the more recent documentation. From a design point of view, a
key difference in approach between \texttt{nls()} and
\texttt{nlsr::nlxb()} is that \texttt{nls()} builds a large
infrastructure, especially the \texttt{nlsModel()} function, from which
the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became Nash
(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
post-solution analysis phases of computations.

@@

Hans Werner Borchers has been helpful in developing the proposal for
this project and in comments on this and related work. Heather Turner
co-mentored the project and helped guide the progress of the work.
Exchanges with Fernando Miguez helped to clarify aspects of selfStart
models and improve package \texttt{nlsr} in that respect.

>>>>

  pdftitle={Refactoring nls(): a constructive critique},

@@

\title{Refactoring \texttt{nls()}: a constructive critique}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2022-6-15}

@@

Important considerations are the usability and maintainability of the
code base that provides the functionality \texttt{nls()} claims to
offer. Our work suggests that the existing code makes maintenance and
improvement very difficult, with legacy applications and examples
blocking some important updates. Discussion of these matters is relevant
to improving R generally as well as its nonlinear estimation tools.

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones which call
  calculations in other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to a number of pre-coded (\texttt{selfStart}) models
\end{itemize}

@@

Adding a small quantity to the loss function before dividing avoids
trouble. In 2021, one of us (J. Nash) proposed that
\texttt{nls.control()} have an additional parameter \texttt{scaleOffset}
with a default value of zero. Setting it to a small number -- 1.0 is a
reasonable choice -- allows small-residual problems (i.e., near-exact
fits) to be dealt with easily. We call this the \textbf{safeguarded
relative offset convergence criterion}. The default value gives the
legacy behaviour. We are pleased to note that this improvement is now in
the R distributed code since version 4.1.0.

@@

The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from
analytic expressions, while that of \texttt{nls()} mostly uses a
numerical approximation. Some \texttt{selfStart} models do provide
``gradient'' information. \texttt{minpack.lm::nlsLM()} invokes
\texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
it appears to use an internal approximate Jacobian code from the
original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such
differences in approach can lead to different behaviour, though in our
experience differences are usually minor.

@@

Note that the \texttt{selfStart} models in the base R
\texttt{./src/library/stats/R/zzModels.R} file and in package
\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
may provide the Jacobian in the ``gradient'' attribute of the
``one-sided'' formula that defines each model, and these Jacobians may
be the analytic forms. The \texttt{nls()} function, after computing the
``right hand side'' or \texttt{rhs} of the residual, then checks to see
if the ``gradient'' attribute is defined, and, if not, uses
\texttt{numericDeriv} to compute a Jacobian into that attribute. This
code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
functions. The use of analytic Jacobians almost certainly contributes to
the good performance of \texttt{nls()} on \texttt{selfStart} models.

@@

Our concern with \texttt{na.action} is that users may be unaware of the
effects of an option setting they may not even be aware has been set.
Should \texttt{na.fail} be the default?

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{\texttt{selfStart}} models. There are
a number of these in base R (\texttt{./src/library/stats/R/zzModels.R})
and others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{\texttt{selfStart}} codes entangle the calculation of starting
values for parameters with the particulars of the \texttt{nls()} code.
Though there is a \texttt{getInitial()} function, this is not easy to
use to simply compute the initial parameter estimates outside of
\texttt{nls()}, in part because it may call that function. Such circular
references are, in our view, dangerous. Moreover, we believe that it
would be helpful to have \texttt{selfStart} models that allow users to
explicitly provide values other than those suggested for the starting
parameters.

Other concerns are revealed by the example below. Here, the
\texttt{SSlogis} \texttt{selfStart} function is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by
\texttt{SSlogis} is
\(y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))\), but we
show how the starting parameters for this model can be transformed to
those of another form of the model, namely,
\(y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))\).

@@

In the event that a \texttt{selfStart} model is not available,
\texttt{nls()} sets all the starting parameters to 1. This is, in our
view, tolerable, but could be improved by using a set of values that are
all slightly different, which, in the case of the example model
\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
avoid a singular Jacobian because \(b\) and \(d\) were equal in value. A
sequence like 1.0, 1.1, 1.2, 1.3 for the four parameters could be
provided quite simply instead of all 1's.

\hypertarget{strategic-issues-in-selfstart-models}{%
\subsubsection{\texorpdfstring{Strategic issues in \texttt{selfStart}
models}{Strategic issues in selfStart models}}\label{strategic-issues-in-selfstart-models}}

By providing starting values that are likely to be reasonable and by
including Jacobian code in the \texttt{selfStart} model code, some of
the deficiencies of the Gauss-Newton algorithm are mitigated. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

Let us compare the \texttt{nls()} result with that from
\texttt{nlsr::nlxb()}, which ostensibly solves the same problem:

@@

The result object from \texttt{nlsr::nlxb()} is still much smaller than
the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
returns the sum of squares as well as the residual vector and Jacobian.
The counts of evaluations are also returned. (Note that the singular
values of the Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result.) As we made our comparisons, we
noted several potential updates to the \texttt{nlsr} documentation as
well as that for \texttt{nls()}.

@@

The variable projection method (Golub and Pereyra
(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
effective than general approaches in finding good solutions to nonlinear
least squares problems when some of the parameters appear linearly.
However, setting up the calculations, that is, identifying which
parameters are linear, is not a trivial task.

@@

The nonlinear modeling specifications are, of course, a development of
linear ones. Unfortunately, the explicit model
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
linear modeling function \texttt{lm()}, which requires this model to be
specified as \texttt{y\ \textasciitilde{}\ x}.

@@

  \DocumentationTok{\#\# Then we use nls\textquotesingle{} indexing feature for parameters in non{-}linear}
  \DocumentationTok{\#\# models to use the conventional algorithm to fit a model in which}
  \DocumentationTok{\#\# alpha and beta vary with Strip.  The starting values are provided}
  \DocumentationTok{\#\# by the previously fitted model.}
  \DocumentationTok{\#\# Note that with indexed parameters, the starting values must be}
  \DocumentationTok{\#\# given in a list (with names):}
\NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
\NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{2}\NormalTok{], nl), }\AttributeTok{b =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{3}\NormalTok{], nl), }\AttributeTok{th =}\NormalTok{ b[}\DecValTok{1}\NormalTok{]))}
  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.2}\NormalTok{)}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }
           \AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Indexed models clearly have a place in some areas of research. We need
to be able to process models such as
\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)},
where \texttt{Strip} is the index. Data \texttt{Length} and
\texttt{Conc} are available for all observations and parameter
\texttt{th} applies to every fitted point, but \texttt{a{[}Strip{]}} and
\texttt{b{[}Strip{]}} apply only to data indexed, which could be an
experimental run number, or some other categorization.

Given that \texttt{{[}\ {]}} provides the R method to index arrays, the
presence of square brackets in a model formula signals indexing. The
question is how to perform the correct calculations.

@@

We anticipate that tools like \texttt{nlsr} will become more attractive
as users discover the increased robustness and general capabilities,
along with the more recent documentation. From a design point of view, a
key difference in approach between \texttt{nls()} and
\texttt{nlsr::nlxb()} is that \texttt{nls()} builds a large
infrastructure, especially the \texttt{nlsModel()} function, from which
the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became Nash
(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
post-solution analysis phases of computations. There is, however, the
wrapper \texttt{wrapnlsr()} that solves the nonlinear estimation problem
with the solver of \texttt{nlsr} then calls \texttt{nls()} to build the
object of class \texttt{nls} for which profile log-likelihood
information can be computed.

@@

Hans Werner Borchers has been helpful in developing the proposal for
this project and in comments on this and related work. Heather Turner
co-mentored the project and helped guide the progress of the work.
Exchanges with Fernando Miguez helped to clarify aspects of
\texttt{selfStart} models and improve package \texttt{nlsr} in that
respect.
